hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-02-28 inst/manuscript/paper.Rmd
edits to case study 2

For the second example, we will again use Bull Trout, but demonstrate a much more complex application for the purpose of showing the full flexibility of the programmatic application of **hatchR**. In this scenario we will use data from @isaak2018 (included `idaho` data set), which includes temperature data from 226 sites across the major upper Columbia River headwater watersheds in Idaho. For this approach we winnow putative bull trout spawning sites by filtering for sites with mean August temperature \</= 13 °C in accordance with thresholds from @isaak2015. For the resulting 139 sites we will demonstrate predicting hatch timing in these putative Bull Trout spawning habitats.

@@

We need to setup our models and data for this analysis, which we don't demonstrate here for the sake of concision, however they are demonstrated in `paper.Rmd` included in the GitHub repository for **hatchR**. After the setup, we can easily map `predict_phenology()` across all putative spawning sites and three spawn dates (September 1-Early Spawning, September 15-Peak Spawning, and September 31-Late Spawning), the results of which are presented in Figure 5.

>>>>

**Bull Trout in the Upper Columbia River Headwaters** OR
**Large-scale Predictions of Bull Trout development timing**

For the second case study, we demonstrate a more complex, large-sclae application of **hatchR**, highlighting its full flexibility when applied programmatically in R. This example also focuses on Bull Trout, but extends beyond a single site to a broad spatial analysis across 226 locations in the upper Columbia River headwaters in Idaho.

We use the `idaho` dataset from @isaak2018, which includes temperature data from these sites. To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature \</= 13 °C, a known thermal threshold for Bull Trout spawning suitability. The filtering process reduced the dataset to 139 sites potential spawning sites.

@@

To predict hatch timing across these sites, we:

1. Set up the necessary moels and data, using Bull Trout parameterization. (For conciseness, we omit this set up here, but full details are available in the `paper.Rmd` file in the GitHub repository for **hatchR**.)

2. Apply `predict_phenology()` across all 139 sites, running predictions for three representative spawn date:

* Early: September 1st
* Peak: September 15th
* Late: September 30th

By mapping predictions across this broad spatial extent, we generate a large-scale assessment of Bull Trout phenology, illustrating how hatch timing caries across different spawning habitats. The results of this analysis are presented in Figure 5, providing insights into how emergence timing might vary under different thermal regimes and across the species' geographic range. This case study underscores the power of **hatchR** for large-scale ecological applications, particularly in conservation planning and habitat suitability assessments.
